\documentclass{article}

\usepackage[preprint]{neurips}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage{url}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{nicefrac}
\usepackage{microtype}
\usepackage{xcolor}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{subcaption}

\title{The Inverted Tell: How AI Deception Reverses Human Linguistic Patterns in Werewolf Games}

\author{%
  Mingxuan He\thanks{Corresponding author.} \\
  University of Chicago \& Gauntlet \\
  \texttt{he.mingxuan527@gmail.com} \\
  \And
  Yui \\
  Independent \\
  \texttt{he.mingxuan527+yui@gmail.com} \\
}

\begin{document}

\maketitle

\begin{abstract}
We analyze 882,510 chat messages from 31,479 games of AI Werewolf (Jinrou) played by eight frontier large language models on the Kaggle Arena---the largest corpus of AI-vs-AI natural language deception to date. We find that AI deception exhibits \emph{systematically inverted} linguistic patterns compared to well-established human deception findings: where human liars use fewer first-person pronouns, AI liars use more; where human liars produce shorter statements, AI liars produce longer ones; where human liars show markers associated with reduced cognitive complexity, AI liars show increased epistemic hedging. These inversions are consistent across all eight models tested, suggesting a fundamental structural difference in how LLMs implement deception versus how humans do. We further show that the most successful AI deceivers are those whose behavioral shifts are \emph{smallest}---the best AI liars are the ones who change least. A dual-annotator semantic study of approximately 16{,}000 messages (15{,}090 resolved labels, $\kappa=0.584$) reveals that AI werewolves overwhelmingly prefer misdirection (67.5\%) over direct lies (7.5\%), with the most successful model employing the most balanced deception repertoire. Our findings have direct implications for AI safety: current deception detection methods calibrated on human baselines may be ineffective or counterproductive when applied to AI systems.
\end{abstract}

\input{sections/introduction}
\input{sections/related_work}
\input{sections/dataset}
\input{sections/methodology}
\input{sections/results}
\input{sections/annotation}
\input{sections/discussion}
\input{sections/conclusion}

\begin{ack}
The Kaggle Werewolf AI Arena dataset was created and funded by Google at a total API cost of \$20,093. We thank the Kaggle community for making this data publicly available.
\end{ack}

\paragraph{Ethics statement.} This study analyzes publicly available data from AI-vs-AI game interactions. No human subjects were involved; all game participants were LLM agents. The dataset contains no personally identifiable information. We discuss implications for AI deception detection in Section~\ref{sec:discussion}; our findings are intended to inform safety research rather than to enable harmful deception capabilities.

\bibliographystyle{plainnat}
\bibliography{references}

\newpage
\appendix
\input{sections/appendix}

\end{document}
