\section{Conclusion}
\label{sec:conclusion}

We presented the first large-scale linguistic analysis of AI deception, analyzing 882,510 chat messages from 31,479 games of AI Werewolf played by eight frontier LLMs. Our central finding is that AI deception \emph{systematically inverts} established human deception patterns: AI liars use more first-person pronouns, write longer messages, and hedge more---the opposite of what decades of human deception research would predict. This inversion is consistent across all eight models tested, spanning four providers and diverse architectures.

We further showed that the most successful AI deceivers are those whose behavior changes least between honest and deceptive roles. This ``best liars change least'' finding suggests that as AI systems become more capable, their deception may become progressively harder to detect through behavioral analysis.

These results challenge the assumption that human deception findings transfer to AI systems and call for the development of AI-specific deception detection methods. We release our analysis code and processed dataset to support future research.\footnote{Code and data: \url{https://github.com/yui-sh/jinrou-paper}}

\paragraph{Future work.} Promising extensions include a direct human comparison using the same game setup, longitudinal analysis of how deception patterns evolve as models are updated, and investigation of whether the deception strategy profiles identified in Section~\ref{sec:annotation} generalize to other social deduction games and adversarial settings.
